Global tradeoffs between prosperity, inequality, and the environment

### 2023-07-04 paper/manuscript.Rmd
changes in response to reviewers

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^, Emilie Lindkvist^2^, Maike Hamann^6^, Jiangxiao Qiu^7^, 
    | \small Caroline Schill^2,8^, Alon Shepon^9^, Anne-Sophie Crépin^8^, and Carl Folke^2,8^
    | \small
    | \footnotesize ^1^The Natural Capital Project and Stanford Center on China's Economy and Institutions, Freeman Spogli 
    | \footnotesize Institute for International Studies, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
    | \footnotesize ^7^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^8^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^9^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}

output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
csl: /Users/juanrocha/Documents/styles/apa.csl
bibliography: references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

@@

- Historical data reveals that no country in the world has reduced inequality and environmental footprint simultanously
- A trilemma exist where countries need to trade-off progress on social and environmental goals
- An inequality trap seems to exist particularly acute in Latinamerica
- Countries trajectories on the trilemma space suggest that in the future social progress can be achieved without compromising the biosphere

@@

For most of human history, material want, not sufficiency, was the norm for most people regardless of county [@Maddison:2007; @Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished[@Alston_20220]). Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and understandably aspires for higher and more secure living standards [@TheWor:2018]. Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an increase rise of global poverty unprecedented since the end of World War II and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].

At the same time, while average incomes for certain countries in what was once called the "Third World" began to  converge with those of the advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
The widespread feelings of injustice that the increasingly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009]. Over the past decade, inequality has once again become an increasingly urgent issue for policymakers as well as the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020]. These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks[@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017]. Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened income and wealth inequalities within many countries, as well as widening the development trajectories between high- and low-income economies [@WorldBank:2022].

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{fig1_trilemma_space}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A). There could be synergies and trade-offs in achieving these three objectives (B-D). We explore trajectory typologies in the trilemma space (E-G). 53 trajectories from 1995 to 2017 are depicted. The trilemma is projected in each of the facets of the tridimensional cube: inequality and ecological footprint (E), inequality and prosperity (F), and ecological footprint and prosperity (G). Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
\label{fig:trilemma}
\end{figure*}
```

@@

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of countries within the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by refs [@Charrad:2014tp; @Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

An analysis of the missing values in the publicly available data shows that assessing changes over time in the trilemma space was only possible for 53 countries. The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $\leqslant 30\%$ of missing values and then imputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

@@

There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). The inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters. Despite this clustering, countries showed distinctive profiles in terms of their direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini and ecological footprint worsened while per-capita GNI grew significantly. And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

@@

In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase in environmental impact with growing prosperity, but this trend reverses after reaching an inflection point of affluence[@barbier1997economic]. As with the original KC, empirical studies of a wide range of variables have come to conflicting conclusions. While some identified international evidence for this inverted-U-shaped relationship, others did not [@dasgupta2002confronting; @stern2004rise; @pothen:2019]. In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we also did not detect an EKC-type trend: i.e., an inflection point in the prosperity-environmental impact plane (Fig \ref{fig:cube_ensamble}). Various mechanisms have been proposed to explain this relationship (or its absence), including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325]. But what is clear from diverse empirical studies is that the connections between prosperity and environmental impact are mediated or confounded by other, largely social and institutional, factors.

@@

Additionally, complex interactions among the three dimensions of the trilemma - and their implications for development - have also been recognized in interdisciplinary studies (none of which posit simple linear or quadratic relationships between the economic, social, and environmental dimensions of sustainability). For instance, Weinzettel et al. [@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions. Similarly, population, environmental carrying capacity, and pollution exhibited mutually-reinforcing feedbacks across countries [@LaTorr:2019]. These types of feedbacks can create and maintain "red loop" or "green loop" dynamics between poverty and the environment, entrapping some countries in long-term conditions of high poverty and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018; @Hamann:2015].

>>>>

---
title: "Global tradeoffs between prosperity, inequality, and the environment"
author: |
    | \small Tong Wu^1,§^, Juan C. Rocha^2,3,§^, Kevin Berry^4^, Tomas Chaigneau^5^,  Maike Hamann^6^, Emilie Lindkvist^2^, Jiangxiao Qiu^7^, 
    | \small Caroline Schill^2,8^, Alon Shepon^9^, Anne-Sophie Crépin^8^, and Carl Folke^2,8^
    | \small
    | \footnotesize ^1^The Natural Capital Project and Institute for International Studies, Stanford University
    | \footnotesize ^2^Stockholm Resilience Centre, Stockholm University
    | \footnotesize ^3^Future Earth
    | \footnotesize ^4^College of Business and Public Policy, University of Alaska Anchorage
    | \footnotesize ^5^Environment and Sustainability Institute, University of Exeter
    | \footnotesize ^6^Center for Complex Systems in Transition, Stellenbosch University
    | \footnotesize ^7^School of Forest Resources & Conservation, Fort Lauderdale Research and Education Center, University of Florida
    | \footnotesize ^8^Beijer Institute of Ecological Economics, The Royal Swedish Academy of Sciences
    | \footnotesize ^9^Department of Environmental Studies, The Porter School of the Environment and Earth Science, Tel Aviv University
    | \footnotesize ^§^These authors made equal contributions
#   | \small \texttt{\href{mailto:juan.rocha@su.se}{\nolinkurl{juan.rocha@su.se}}}
output:
  pdf_document:
    dev: pdf
    keep_tex: yes
    toc: no
    latex_engine: xelatex
  rtf_document:
    toc: no
  word_document:
csl: /Users/juanrocha/Documents/styles/apa.csl
bibliography: references_ei.bib
citecolor: blue
documentclass: article
font: Arial
fontsize: 10pt
linkcolor: blue
urlcolor: blue
header-includes:
- \usepackage{dcolumn, rotating, longtable, lineno, float, array, tabularx, inputenc}
- \setmainfont{Arial}
- \renewcommand{\abstractname}{}
- \graphicspath{{figures/}}
- \usepackage[margin=2.5cm]{geometry}
- \linenumbers
- \usepackage{setspace}
---

@@

- Historical data reveals that no country in the world simultaneously achieved high prosperity, low inequality, and low environmental impact.
- A sustainability trilemma seems to exist where progress on either social goal comes at the expense of the environment.
- An inequality trap seems to be particularly acute in Latin America.
- Countries trajectories on the trilemma space suggest that future social progress can be achieved without compromising the biosphere.

@@

For most of human history, material want, not sufficiency, was the norm for most people regardless of country [@Maddison:2007; @Harari:2014]. Although global prosperity has increased in recent generations, progress has been uneven. If China’s gains in poverty alleviation were removed, the United Nations’ Millennium Development Goal of halving extreme poverty between 1990 and 2015 would not have been achieved [@ki2013millennium]. And even today, a significant share of the world’s population still lives under the international poverty line (a standard that the UN itself recognizes may grossly undercount the world's impoverished [@Alston_20220]). Right above this derisory threshold,  hundreds of millions more people maintain a precarious subsistence and understandably aspire for higher and more secure living standards [@TheWor:2018]. Furthermore, the recent COVID-19 pandemic pushed tens-of-millions of people back beneath that already low poverty line, leading to an increase of global poverty unprecedented since the end of World War II and underscoring the economic precarity that still afflicts much of the world [@buheji2020extent].

At the same time, while average incomes for certain countries in what was once called the "Third World" began to  converge with those of the advanced economies in recent decades, asymmetries in the domestic distribution of income and wealth have worsened within both, developed and developing countries [@Bourguignon:2007; @Milanovic:2016].
The widespread feelings of injustice that the increasingly unequal economic distribution arouses has historically been called the "Social Question", inspiring influential theories of political transformation in the 19th and 20th Centuries (for instance various strands of Marxist and Third-Worldist thought) [@Christiansen:2019; @Faist:2009]. Over the past decade, inequality has once again become an increasingly urgent issue for policymakers as well as the general public, prompting anxieties about social and political instability [@Chancel:2020; @UN2020]. These anxieties have been reinforced by a growing body of research demonstrating complex linkages between inequality and a range of financial, health, and ecological risks [@Hamann:2018; @Leach:2018; @Pikett:2014; @Wilkinson:2009; @Hartmann:2017]. Furthermore, the impacts and policy responses from the COVID-19 pandemic worsened income and wealth inequalities within many countries, as well as widened the development trajectories between high- and low-income economies [@WorldBank:2022].

@@

```{=tex}
\begin{figure*}[ht]
\centering
\includegraphics[width = 6in, height = 4in]{fig1_trilemma_space}
\caption{\textbf{Trilemma space} Countries around the world face the challenge of reducing inequality and environmental impact while increasing prosperity (A). There could be synergies and trade-offs in achieving these three objectives (B-D). We explore trajectory typologies in the trilemma space (E-G). 53 country trajectories from 1995 to 2017 are depicted. The trilemma is projected in each of the facets of the tridimensional cube: ecological footprint and prosperity (E), inequality and prosperity (F), and inequality and ecological footprint (G). Thresholds that delimit the desirable space are depicted as dashed lines and the colours represent negative synergies (red), positive synergies (blue), and tradeoffs (light orange) between the dimensions of the trilemma (see Methods).}
\label{fig:trilemma}
\end{figure*}
```

@@

We used these data to chart the movement of countries across the dimensions of prosperity, inequality, and environmental impact over time. We further investigated whether these movements trace out typologies of country trajectories within the trilemma space. To identify different national profiles, we calculated the magnitudes and angles of displacement of countries over time.  We then used average and standard deviation statistics to discover country clusters and performed sensitivity analyses to determine the relevant number of clusters (clustering is a statistical process whereby objects are partitioned and then grouped in such a way that the members of one cluster are more closely related to each other than they are to those of other clusters). We compared ten clustering algorithms over 30 different metrics of performance, following protocols outlined by Charrad _et al._, [-@Charrad:2014tp] and Brook _et al._ [-@Brock:2008vz]. We used principal component analysis as an ordination method for visualization, and then vector-fitting to show how clusters are pulled apart in the principal component space (in order to make the three-dimensional distribution of countries intelligible). All data used are publicly available, and the code used for the analysis is available at: <link will be added upon publication, avoided here due to double-blind requirements>.

An analysis of the missing values in the publicly available data shows that assessing changes over time in the trilemma space was only possible for 53 countries. The most incomplete dataset is that for inequality (Figs \ref{fig:SM_gni_missing}, \ref{fig:SM_inq_missing}, \ref{fig:SM_ecofp_missing}), where for any given year at least half of the countries of the world have missing values. Based on the quality of the three datasets, we selected time series from 1995-2017 for the 53 countries with $ fewer than 30\% of missing values and then imputed them with a linear interpolation. This reduced dataset, although incomplete, enables the study of the dynamics of countries in the trilemma space (Fig \ref{fig:trilemma}).

@@

There are certain regions of the trilemma space where countries tended to dwell (Fig \ref{fig:cube_ensamble}). The prosperity-ecological footprint plane (Fig \ref{fig:cube_ensamble}A), and inequality-ecological footprint plane (Fig \ref{fig:cube_ensamble}C) have two of these attractors, while the inequality-prosperity plane (Fig \ref{fig:cube_ensamble}B) has three, suggesting the existence of two to three country trajectory clusters. Despite this clustering, countries showed distinctive profiles in terms of their direction of movement (Fig \ref{fig:cube_ensamble} insets). Most countries had low displacement (short arrows), meaning that their trajectory over the study period did not take them far from the region in the trilemma space where they started. A few countries were distinctive in their direction and magnitude of displacement from their starting point in 1995. For example, between 1995 and 2017, China's Gini and ecological footprint worsened while per-capita GNI grew significantly. And while inequality and environmental impact actually fell in Ukraine, prosperity only increased marginally. To better understand country typologies, we then clustered summary statistics (mean and standard deviations) of the three trilemma dimensions, as well as the direction and magnitude of country movement over time (Fig \ref{fig:pca}).

@@

In terms of the economy-environment relationship, the Environmental Kuznets Curve (EKC) has often been advanced as an axiomatic relationship: countries increase in environmental impact with growing prosperity, but this trend reverses after reaching an inflection point of affluence [@barbier1997economic]. As with the original KC, empirical studies of a wide range of variables have come to conflicting conclusions. While some identified international evidence for this inverted-U-shaped relationship, others did not [@dasgupta2002confronting; @stern2004rise; @pothen:2019]. In the trilemma space, with the composite indicator of ecological footprint as the environmental indicator, we also did not detect an EKC-type trend: i.e., an inflection point in the prosperity-environmental impact plane (Fig \ref{fig:cube_ensamble}). Various mechanisms have been proposed to explain this relationship (or its absence), including market liberalization, more stringent regulations, and greater freedoms for citizens to voice their concerns [@MALER201325]. But what is clear from diverse empirical studies is that the connections between prosperity and environmental impact are mediated or confounded by other, largely social and institutional, factors.

@@

Additionally, complex interactions among the three dimensions of the trilemma - and their implications for development - have also been recognized in interdisciplinary studies (none of which posit simple linear or quadratic relationships between the economic, social, and environmental dimensions of sustainability). For instance, Weinzettel et al. [-@Weinzett:2013] found that countries with high per-capita biocapacity (i.e., large natural resource endowments relative to population level) tended to spare more land for nature, forming a virtuous cycle of improving environmental conditions. Similarly, population, environmental carrying capacity, and pollution exhibited mutually-reinforcing feedbacks across countries [@LaTorr:2019]. These types of feedbacks can create and maintain "red loop" or "green loop" dynamics between poverty and the environment, entrapping some countries in long-term conditions of high poverty and environmental degradation while placing others on a more sustainable trajectory [@Cumming:2014; @Cumming:2018; @Hamann:2015].
